Class Project Report: Boundary-free Technology Mapping using Gr\"obner Basis

### 2015-05-01 techwriting/ASPDAC/theory.tex
all write-up from virtue machine

\section{Essential Theory}
\subsection{Polynomial Representation}
A typical sequential circuit is composed as figure \ref{fig:sequential}. The combinational logic has primary input $x$, state
inputs $\{s_0, s_1, \dots, s_{k-1}\}$, primary output $z$ and state outputs $t_0, t_1, \dots, t_{k-1}$. State outputs
are latched into registers, which will be fed back to state inputs in next clock cycle.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=3.5in]{./sequential_fig.eps}

\end{center}
\caption{A typical sequential circuit block view}
\label{fig:sequential}
\end{figure}

A transition function $\Delta$ is a Boolean function about state outputs and all inputs, i.e.
\begin{displaymath}
t_i = \Delta_i(x, s_0, s_1, \dots, s_{k-1})
\end{displaymath}
This constrain could also be written in the form of equation $polynomial = 0$:
\begin{displaymath}
t_i - \Delta_i(x, s_0, s_1, \dots, s_{k-1}) = 0
\end{displaymath}
Solution to this equation is limited within $\{0, 1\}$, which is set of all elements in Galois field $\mathbb{F}_2$.
Furthermore, there exists a one-to-one corresponding relation from Boolean space to Galois field on higher dimension: 
$\mathbb{B}^k \to \mathbb{F}_{2^k}$, which means it is possible to find a set of elements in $\mathbb{F}_{2^k}$ representing
all Boolean values every bit of a bit vector can take. In Galois field, an example is in table \ref{table:booltogalois}.
\begin{table}[tb]
\centering
\caption{Bit-vector, Exponential and Polynomial representation of
elements in  ${\mathbb{F}}_{2^4} = {\mathbb{F}}_2[x]
\pmod{x^4+x^3+1}$}
\begin{tabular}{|c|c||c|c|} 
\hline
$a_3a_2a_1a_0$ & Polynomial     &$a_3a_2a_1a_0$ & Polynomial  \\
\hline
$0000$        & $0$           & $1000$  &$\alpha^3$\\
\hline
$0001$        & $1$           & $1001$  & $\alpha^3 + 1$\\
\hline
$0010$        &  $\alpha$       & $1010$ & $\alpha^3 + \alpha$  \\
\hline
$0011$        &  $\alpha + 1$   & $1011$ &  $\alpha^3+\alpha+1$\\
\hline
$0100$        &  $\alpha^2$     &  $1100$ &  $\alpha^3 + \alpha^2$\\
\hline
$0101$        & $\alpha^2 + 1$ & $1101$  & $\alpha^3+\alpha^2+1$\\
\hline
$0110$        &  $\alpha^2 + \alpha$ & $1110$ &  $\alpha^3+\alpha^2+\alpha$\\
\hline
$0111$        & $\alpha^2+\alpha+1$ & $1111$ & $\alpha^3+\alpha^2+\alpha+1$\\
\hline
\end{tabular}
\label{table:booltogalois}  
\end{table}
A word-level representation on state inputs/outputs can be attained through above correspondence. If the set of 
state inputs $\{s_0, s_1, \dots, s_{k-1}\}$ is denoted by $S$ taking values from $\mathbb{F}_{2^k}$, and
set of outputs $\{t_0, t_1, \dots, t_{k-1}\}$ is denoted by $T$, a word-level transition function can be written as
\begin{displaymath}
T - \Delta(x, S) = 0
\end{displaymath}
If $T - \Delta(x, S)$ is taken as a polynomial $f$, then $f \in \mathbb{F}_{2^k}[x]\ mod\ P(\alpha)$.
\subsection{Gr\"obner Basis Theory}
Gr\"obner basis is used to calculate desired results out of a set of polynomials.
If there exists multiple polynomials $\{f_1, f_2, \dots, f_r\}$ representing constrains on $T, x$ and $S$,
then all polynomials should simultaneously equal to $0$.
\begin{displaymath}
\left\{
  \begin{array}{lc}
  f_1 & = 0\\
  f_2 & = 0\\
  \dots & \ \\
  f_r & = 0
  \end{array} \right.
\end{displaymath}
Solution to these equations is set of values of $(T, x, S)$, which is called \textbf{Variety} of \textbf{ideal}
$I = \langle f_1, f_2, \dots, f_r\rangle $.

Calculate Reduced Gr\"obner Basis (RGB) from ideal $I$ with \textbf{elimination term order}, there exists a polynomial contains
only $T$, if the values of states input $S$ and primary input $x$ are given in the form of polynomials and included by $I$,
the values of $T$ can be computed.

Otherwise, if RGB calculation is manipulated under \textbf{abstraction term order}, there exists a polynomial in
the form of $T - \mathcal{F}(x, S)$ representing transition function. In next clock cycle, assign new state input $S'$ with $T$, the next
state can be computed again, result is polynomial $T' - \mathcal{F}(x', S')$.

@@

\begin{figure}[hbt]
\begin{center}
\includegraphics[scale=0.3]{./stg_fig.eps}

\end{center}
\caption{State transition graph of sample FSM}
\label{fig:stg}
\end{figure}

\begin{figure}[hbt]
\begin{center}
\includegraphics[width=3.5in]{./fsm_fig.eps}

\end{center}
\caption{Gate-level circuit of combinational logic block of sample FSM}
\label{fig:fsm}
\end{figure}

The first example is a mealy finite state machine (FSM). Gate-level circuits of combinational logic
is figure \ref{fig:fsm}, $x$ is primary input,
$\{s_0, s_1\}$ are state inputs, $Z$ is primary output and $\{t_0, t_1\}$ are state outputs.
Word-level state input is $S = s_0 + s_1\cdot\alpha$, output is $T = t_0 + t_1\cdot\alpha$. In Galois field
$\mathbb{F}_{2^2}$, bit-level variables $x, s_0, s_1, Z, t_0, t_1$ can take values from $\{0, 1\}$, while
word-level variables $S, T$ may take values from $\{0, 1, \alpha, 1 + \alpha\}$. State transition graph (STG)
showed in figure \ref{fig:stg} uses 2-bit Boolean vector to represent 4 states $\{S_0, S_1, S_2, S_3\}$, which
could be converted to elements in $\mathbb{F}_{2^2}$ similarly as table \ref{table:booltogalois} shows.

One important technique to check sequential equivalence is \textbf{state space traversal}. In this example,
Breath-First-Search (BFS) is employed for traversal, which can be described with following algorithm\cite{KallaPartialScan}:

@@

\begin{figure*}[tb]
\begin{center}
\includegraphics[width=\textwidth]{mySMPO.eps}

\end{center}
\caption{5-bit SMPO. Index $i$ satisfies $0<i<4$, indices $u,v$ are determined by column \# of nonzero entries in $i$-th row of $\lambda$-Matrix $M^{(0)}$, i.e. if entry $M_{ij}^{(0)}$ is a nonzero entry, $u$ or $v$ equals to $i+j \pmod 5$. Index $w = 2i\pmod 5$}
\label{fig:SMPO}
\end{figure*}

@@

\begin{figure}[hbt]
\begin{center}
\includegraphics[width=3.5in]{./RH.eps}

\end{center}
\caption{5-bit RH-multiplier (AESMPO)}
\label{fig:RHmulti}
\end{figure}

@@

Apply this approach to calculate image in BFS traversal, but modify algorithm
\ref{alg:BFS} to make it adapt 5 steps reachable states enumeration.
The result is $R + AB$, which validates the function of this circuit.

@@

\begin{Example}
{\bf Objective}:\ Abstract polynomial $a_i + \mathcal{G}_i(A)$ from $f_0: a_0\alpha^5+a_1\alpha^{10}+a_2\alpha^{20}+a_3\alpha^9+a_4\alpha^{18}+A$.

First, compute $f_0^2: a_0\alpha^{10}+a_1\alpha^{20}+a_2\alpha^{9}+a_3\alpha^{18}+a_4\alpha^{5}+A^2$. Apparently variable $a_0$ can be
eliminated by operation 
\begin{align}
f_1 =& f_0\times \alpha^5 + f_0^2: \nonumber\\
&a_1+(\alpha)a_2+(\alpha^4+\alpha^2)a_3+(\alpha^3+\alpha^2)a_4\nonumber\\
+&(\alpha^4+\alpha^3+\alpha^2+1)A^2+(\alpha^2+\alpha)A\nonumber
\end{align}
Recursively eliminate $a_1$ from $f_1$, $a_2$ from $f_2$, etc. The final polynomial $f_4$ has the form 
$a_4 + \mathcal{G}_4(A) = a_4+(\alpha^4+\alpha^3+\alpha)A^{16}+(\alpha^3+\alpha^2)A^8+(\alpha^4+1)A^4+(\alpha^2+1)A^2+(\alpha+1)A$. Recursively substitute
$\mathcal{G}_4(A)$ back to $f_3$, etc. The result is a set of polynomials
\begin{displaymath}
\{g_i\ | \ g_i: a_i + \mathcal{G}_i(A)\}
\end{displaymath}
\end{Example}
In this approach it is easy to get word-level variable representation for each bit-level primary inputs. By substitution, a new polynomial in the form $R+\mathcal{F}(A,B)$
could be attained.

>>>>

\section{Essential Theory}
\subsection{Polynomial Representation}
A typical sequential circuit model is showed as blocks in figure \ref{fig:sequential}. The combinational logic block 
has primary inputs $X = \{x_0, x_1, \dots, x_{m-1}\}$, state
inputs $S = \{s_0, s_1, \dots, s_{k-1}\}$, primary output $Z$ and state outputs $T = \{t_0, t_1, \dots, t_{k-1}\}$. State outputs
are latched into registers, which will be fed back to state inputs in next clock cycle.

%% \begin{figure}[tb]
%% \begin{center}
%% \includegraphics[width=3.5in]{./sequential_fig.eps}
%% \caption{A typical sequential circuit block view}
%% \end{center}
%% \label{fig:sequential}
%% \end{figure}

A transition function $\Delta$ is a Boolean function giving state outputs from state inputs and primary inputs, i.e.
\begin{displaymath}
t_i = \Delta_i(x, s_0, s_1, \dots, s_{k-1})
\end{displaymath}
This constrain could also be written in the form of equation $polynomial = 0$:
\begin{displaymath}
t_i - \Delta_i(x, s_0, s_1, \dots, s_{k-1}) = 0
\end{displaymath}
Solution to this equation is limited within $\{0, 1\}$, which is set of all elements in Galois field $\mathbb{F}_2$.
Furthermore, there exists a one-to-one corresponding relation from Boolean space to Galois field on higher dimension: 
$\mathbb{B}^k \to \mathbb{F}_{2^k}$, which means it is possible to find a set of elements in $\mathbb{F}_{2^k}$ representing
all Boolean values that a $k$-bit bit-vector can take. In Galois
field, a as \ref{table:booltogalois} shows.

\begin{table}[htb]
\centering
\caption{Bit-vector, Exponential and Polynomial representation of
elements in  ${\mathbb{F}}_{2^4} = {\mathbb{F}}_2[x]
\pmod{x^4+x^3+1}$}
\begin{tabular}{|c|c||c|c|} 
\hline
$a_3a_2a_1a_0$ & Polynomial     &$a_3a_2a_1a_0$ & Polynomial  \\
\hline
$0000$        & $0$           & $1000$  &$\alpha^3$\\
\hline
$0001$        & $1$           & $1001$  & $\alpha^3 + 1$\\
\hline
$0010$        &  $\alpha$       & $1010$ & $\alpha^3 + \alpha$  \\
\hline
$0011$        &  $\alpha + 1$   & $1011$ &  $\alpha^3+\alpha+1$\\
\hline
$0100$        &  $\alpha^2$     &  $1100$ &  $\alpha^3 + \alpha^2$\\
\hline
$0101$        & $\alpha^2 + 1$ & $1101$  & $\alpha^3+\alpha^2+1$\\
\hline
$0110$        &  $\alpha^2 + \alpha$ & $1110$ &  $\alpha^3+\alpha^2+\alpha$\\
\hline
$0111$        & $\alpha^2+\alpha+1$ & $1111$ & $\alpha^3+\alpha^2+\alpha+1$\\
\hline
\end{tabular}
\label{table:booltogalois}  
\end{table}
A word-level representation of state inputs/outputs can be obtained through above correspondence. If the set of 
state inputs $\{s_0, s_1, \dots, s_{k-1}\}$ is denoted by $S$ taking values from $\mathbb{F}_{2^k}$, $X$ stands
for a word taking values from $\mathbb F_{2^m}$ and
the set of outputs $\{t_0, t_1, \dots, t_{k-1}\}$ is denoted by $T$, a word-level transition function can be written as
\begin{displaymath}
T - \Delta(X, S) = 0
\end{displaymath}
Let polynomial $f = T - \Delta(x, S)$, then $f \in \mathbb{F}_{2^k}[x]\ mod\ P(\alpha)$.
\subsection{Gr\"obner Basis Theory}
Gr\"obner basis is used to calculate desired results out of a set of polynomials.
If there exists multiple polynomials $\{f_1, f_2, \dots, f_r\}$ representing constrains on $T, x$ and $S$,
then all polynomials should simultaneously equal to $0$.
\begin{displaymath}
\left\{
  \begin{array}{lc}
  f_1 & = 0\\
  f_2 & = 0\\
  \dots & \ \\
  f_r & = 0
  \end{array} \right.
\end{displaymath}
Solution to these equations is set of values of $(T, x, S)$, which is called \textbf{Variety} of \textbf{ideal}
$I = \langle f_1, f_2, \dots, f_r\rangle $.

Calculate Reduced Gr\"obner Basis (redGB) from ideal $I$ with \textbf{elimination term order}, there exists a polynomial contains
only $T$, if the values of states input $S$ and primary input $x$ are given in the form of polynomials and included by $I$,
the values of $T$ can be computed.

Otherwise, if redGB calculation is manipulated under \textbf{abstraction term order}, there exists a polynomial in
the form of $T - \mathcal{F}(x, S)$ representing transition function. In next clock cycle, assign new state input $S'$ with $T$, the next
state can be computed again, result is polynomial $T' - \mathcal{F}(x', S')$.

@@

\begin{figure}[hbt]
\begin{center}
\includegraphics[scale=0.3]{./stg_fig.eps}
\caption{State transition graph of sample FSM}
\end{center}
\label{fig:stg}
\end{figure}

\begin{figure}[hbt]
\begin{center}
\includegraphics[width=3.5in]{./fsm_fig.eps}
\caption{Gate-level circuit of combinational logic block of sample FSM}
\end{center}
\label{fig:fsm}
\end{figure}

The first example is a mealy Finite State Machine (FSM). Gate-level circuits of combinational logic
is figure \ref{fig:fsm}, $x$ is primary input (one bit),
$\{s_0, s_1\}$ are state inputs, $Z$ is primary output and $\{t_0, t_1\}$ are state outputs.
Word-level state input is $S = s_0 + s_1\cdot\alpha$, output is $T = t_0 + t_1\cdot\alpha$. In Galois field
$\mathbb{F}_{2^2}$, bit-level variables $x, s_0, s_1, Z, t_0, t_1$ can take values from $\{0, 1\}$, while
word-level variables $S, T$ may take values from $\{0, 1, \alpha, 1 + \alpha\}$. State Transition Graph (STG)
showed in figure \ref{fig:stg} uses 2-bit Boolean vector to represent 4 states $\{S_0, S_1, S_2, S_3\}$, which
could be mapped to elements in $\mathbb{F}_{2^2}$ similarly as table \ref{table:booltogalois} shows.

One important technique to check sequential equivalence is \textbf{state space traversal}. In this example,
Breath-First-Search (BFS) is employed for traversal procedure, which can be described by following algorithm\cite{KallaPartialScan}:

@@

\begin{figure*}[tb]
\begin{center}
\includegraphics[width=\textwidth]{mySMPO.eps}
\caption{5-bit SMPO. Index $i$ satisfies $0<i<4$, indices $u,v$ are determined by column \# of nonzero entries in $i$-th row of $\lambda$-Matrix $M^{(0)}$, i.e. if entry $M_{ij}^{(0)}$ is a nonzero entry, $u$ or $v$ equals to $i+j \pmod 5$. Index $w = 2i\pmod 5$}
\end{center}
\label{fig:SMPO}
\end{figure*}

@@

\begin{figure}[hbt]
\centering
\begin{minipage}{12cm}
\includegraphics[width=4.5in]{./RH.eps}
\caption{5-bit RH-multiplier (AESMPO)}
\end{minipage}
\label{fig:RHmulti}
\end{figure}

@@

If we apply this approach to calculate "image function" in BFS traversal, but modify algorithm
\ref{alg:BFS} to make it adapt 5 steps reachable states enumeration,
the result will be $R + AB$ which validates the function of this circuit.

@@

\begin{Example}
{\bf Objective}:\ Abstract polynomial $a_i + \mathcal{G}_i(A)$ from $f_0 = a_0\alpha^5+a_1\alpha^{10}+a_2\alpha^{20}+a_3\alpha^9+a_4\alpha^{18}+A
 = g_0 + A$.

First, compute $f_0^2= a_0\alpha^{10}+a_1\alpha^{20}+a_2\alpha^{9}+a_3\alpha^{18}+a_4\alpha^{5}+A^2 = g_0^2 + A^2$;
then $f_0^4, f_0^8, f_0^{16}$. By repeating squaring we get a system of equations:
\begin{displaymath}
  \begin{cases}
  f_0  &=\ \  0\\
  f_0^2 & =\ \  0\\
  f_0^4  &=\ \  0\\
  f_0^8  &=\ \  0\\
  f_0^{16} & =\ \ 0
  \end{cases}
\Longleftrightarrow\ \
\begin{cases}
  g_0 & =\ \  A\\
  g_0^2 & =\ \  A^2\\
  g_0^4 & =\ \  A^4\\
  g_0^8 & =\ \  A^8\\
  g_0^{16} & =\ \  A^{16}
  \end{cases}
\end{displaymath}
Following is the coefficients matrix form of this system of equations:
\begin{displaymath}
\begin{pmatrix}
\alpha^5 & \alpha^{10} & \alpha^{20} & \alpha^9 & \alpha^{18} \\
\alpha^{10} & \alpha^{20} & \alpha^9 & \alpha^{18} & \alpha^5 \\
\alpha^{20} & \alpha^9 & \alpha^{18} & \alpha^5 & \alpha^{10} \\
\alpha^9 & \alpha^{18} & \alpha^5 & \alpha^{10} & \alpha^{20} \\
\alpha^{18} & \alpha^5 & \alpha^{10} & \alpha^{20} & \alpha^9 
\end{pmatrix}
\begin{pmatrix}
a_0\\
a_1\\
a_2\\
a_3\\
a_4
\end{pmatrix}
=
\begin{pmatrix}
A\\
A^2\\
A^4\\
A^8\\
A^{16}
\end{pmatrix}
\end{displaymath}

Use Gaussian elimination on coefficients matrix, for example  
\begin{align}
Row\ 2 =& Row\ 1\times \alpha^5 + Row\ 2: \nonumber\\
&a_1+(\alpha)a_2+(\alpha^4+\alpha^2)a_3+(\alpha^3+\alpha^2)a_4\nonumber\\
=&(\alpha^4+\alpha^3+\alpha^2+1)A^2+(\alpha^2+\alpha)A\nonumber
\end{align}
Recursively eliminate $a_1$ from third row, $a_2$ from fourth row, etc. The final solution to this system of
equations is
%\begin{displaymath}
%\{g_i\ | \ g_i: a_i + \mathcal{G}_i(A)\}
%\end{displaymath}
%\end{Example}
\begin{displaymath}
\left\{
  \begin{array}{lcl}
  a_0 & = & (\alpha+1)A^{16}+(\alpha^4+\alpha^3+\alpha)A^8+(\alpha^3+\alpha^2)A^4\\&&+(\alpha^4+1)A^2+(\alpha^2+1)A\\
  a_1 & = & (\alpha^2+1)A^{16}+(\alpha+1)A^8+(\alpha^4+\alpha^3+\alpha)A^4\\&&+(\alpha^3+\alpha^2)A^2+(\alpha^4+1)A\\
  a_2 & = & (\alpha^4+1)A^{16}+(\alpha^2+1)A^8+(\alpha+1)A^4\\&&+(\alpha^4+\alpha^3+\alpha)A^2+(\alpha^3+\alpha^2)A\\
  a_3 & = & (\alpha^3+\alpha^2)A^{16}+(\alpha^4+1)A^8+(\alpha^2+1)A^4\\&&+(\alpha+1)A^2+(\alpha^4+\alpha^3+\alpha)A\\
  a_4 & = & (\alpha^4+\alpha^3+\alpha)A^{16}+(\alpha^3+\alpha^2)A^8+(\alpha^4+1)A^4\\&&+(\alpha^2+1)A^2+(\alpha+1)A
  \end{array} \right.
\end{displaymath}
\end{Example}
In this approach it is easy to get word-level variable representation for each bit-level primary inputs. By substitution, a new polynomial in the form $R+\mathcal{F}(A,B)$
could be attained.
